---
title: "Activities - Scatterplot, Correlation, Crime"
author: "Wells"
date: "2025-10-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Activity #1:

Examine this New York Times scatterplot from 2013:
https://archive.nytimes.com/www.nytimes.com/interactive/2013/05/07/education/college-admissions-gap.html?_r=0

The graphic displays four different scatterplots showing relationships between student demographics and college acceptance rates. For each of the four scatterplots:

a) Describe the relationship (or lack thereof) between the two variables.

ANSWER: The first graph takes a snapshot of the second, and shows the percentage of black students that attend and graduate American universities that are defined as "elite." There's very little correlation, but what correlation there is, is positive. Expanding into graph two, there is definitely some negative correlation but there is also a collection of outliers in the upper left portion of the graph. Graph three is similar to two in the sense it displays negative moderate corrleation, but again, there are additional outlier toward the top of the data set. The final graph depicts the schools that Asian studnet typically attend, and their graduation rates at those schools. Through that graph, we see that some light positive correlation, showing that Asian students are more likely to attend and graduate from more elite universities, as compared to other minority groups. 

b) Note the specific words and phrases the graphic uses to characterize each relationship.

ANSWER: Most of the specific key words and phrases for each graph come in the small descriptions in the top right. For the first graph, the comparison to high school graduates is significant. The description also includes a mention of "affirmative action" and "elite colleges." The description for graph two notes that black students are half as likely as white students to recieve a degree, even though the two groups have similar overall rates for attending college. Graph three explains that Hispanic students are more likely to attend community colleges, which "tend to have more dropouts." The final graph shows how Asian students make up a relatively large portion of the graduating class at elite universities, despite making up just "5 percent of public high school graduates."   

Write 1-2 sentences summarizing each scatterplot.

ANSWER: Adressed in my previous answer for part B

Name each person in your group who worked on this entry.

**Write Answer #1:**
Aidan Currie, Ben Messinger, Luke Tabacco 
-------------------------------------------------------------------------------

Activity #2: Correlation Demonstration

In the following exercise, we will look at data about 16 different consultants, their aptitude scores and their performance scores.

Our question: what is the relationship between a consultant's aptitude and their performance? 

How are they correlated?

Below we calculate a Pearson's correlation coefficient and then create a regression line.

```{r}
library(tidyverse)
```

Load data for aptitude and performance for 16 different consultants
```{r}
aptitude <- c(45, 81, 65, 87, 68, 91, 77, 61, 55, 66, 82, 93, 76, 83, 61, 74)
performance <- c(56, 74, 56, 81, 75, 84, 68, 52, 57, 82, 73, 90, 67, 79, 70, 66)
consultant <- c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16)
#this builds our table
consultants <- data.frame(consultant, aptitude, performance)
consultants
```


### Task 1: Use `summarize()` and the `cor()` function to calculate the correlation between aptitude and performance, using the pearson method.

```{r}
consultants |> summarize(correlation = cor(aptitude, performance, method="pearson"))
```

We see a Pearson's correlation coefficient of 0.783.
What does that mean?

The Pearson correlation scale is from -1 to 1.
-1 is a perfect negative correlation
1 is a perfect positive correlation

How do we interpret the values in between?

-- High Degree: Values between ±0.50 and ±1 suggest a strong correlation.
-- Moderate Degree: Values between ±0.30 and ±0.49 indicate a moderate correlation.
-- Low Degree: Values below +0.29 are considered a weak correlation.
-- No Correlation: A value of zero implies no relationship.

**Question: How would you describe the 0.783 correlation between aptitute and performance?**

**A: Write your answer:**

ANSWER: Based on the descriptions above, there is a relatively strong, positive correlation between aptitude and performance. But there's still potential for the correlation to be stronger.

### Task 2: Create a scatterplot of aptitude and performance, using `geom_point()`

```{r}

consultants |> 
  ggplot() +
  geom_point(aes(x=aptitude, y=performance)) +
  geom_smooth(aes(x=aptitude, y=performance), method="lm") +
  expand_limits(x = 0, y = 0)

```


The "best fit" line is known as a regression line that helps us understand the relationship between two variables. This predicts the dependent variable based on the independent variable. 

A regression line is a straight line showing a best-fit between independent and dependent variables.

In the example above, aptitude would be the independent variable, or the predictor. That's because we would expect aptitude to predict performance, which is the dependent variable, or the outcome variable.

Independent = predictor
Dependent = outcome

Regression is one step ahead of correlation. It helps predict the values of dependent variables or outcomes with respect to changes in independent or explanatory variables. 

Regression lines help predict outcomes.


-------------------------------------------------------------------------------

Activity #3: 

# PG Murders and Heat

Let's explore the relationship between crime and hot weather using homicide and temperature data from Prince George's County. 

You'll load the data, perform some basic statistical analysis, and create a scatterplot to explore the relationship. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

**write code below**
```{r}
library(tidyverse)
```


### Task 2: Load the dataset from [https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_murder_temps.csv](https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_murder_temps.csv)

**write code below**
```{r}
crime_to_weather <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_murder_temps.csv")
```



### Task 3: Use `summarize()` and the `cor()` function to calculate the correlation between murders and temperature, using the pearson method.

**write code below**
```{r}
crime_to_weather |> summarize(correlation = cor(murders, temperature, method="pearson"))
```



### Task 4: Create a scatterplot of murders and temperature, using `geom_point()`. Your aes() will have temperatures on the x axis, murders on the y axis and add a label for the month (so label=month inside the aes()).

**write code below**
```{r}
crime_to_weather |> 
  ggplot() +
  geom_point(aes(x=temperature, y=murders)) +
  geom_smooth(aes(x=temperature, y=murders), method="lm", label = month) +
  expand_limits(x = 0, y = 0)
  labs(title = "Correlation Murders and Average Temperature", x ="Average Temperature", y ="Murders")
```


### Task 5: Add labels to the scatterplot using geom_text(). Inside geom_text() put the same aes() as geom_point in Task 4, then after the aes() add size = 5.

**write code below**
```{r}
crime_to_weather |> 
  ggplot() +
  geom_point(aes(x=temperature, y=murders, size = 5)) +
  geom_smooth(aes(x=temperature, y=murders), method="lm") +
  geom_text(aes(x = temperature, y = murders, label = month)) + 
  labs(title = "Correlation Murders and Average Temperature", subtitle = "What's the relationship?", x ="Average Temperature", y ="Murders")
```


### Task 6: Describe the results of both the correlation calculation and the scatterplot.

**Write Answer:**

The correlation coefficient of 0.58 indicates a moderately positive correlation between average temperature and total murders, while the scatterplot gives a more specific depiction of the month-to-month breakdown. Based on the scatterplot, January and September are the only two months that fall closely along the line of best fit, with several other months being in the general range as well. With that being said, months like November and April represent significant outliers. Nevertheless though, there's some correlation between higher temperatures and more murders. 

When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
