---
title: "Survey Weighting Exercise: Nonvoter Study Analysis"
date: "2025-08-21"
author: "Derek Willis/Rob Wells"
---

**YOUR NAME HERE:**
Aidan Currie


Activity: Belonging Survey Methods

Let's revisit the summary of UMD's Belonging Survey, and look at the appendix in particular. https://president.umd.edu/internal/belonging

Working in pairs or small groups, each person reads the appendix closely and identifies any statements that raise methodological concerns or are unclear. Discuss your findings, summarize your discussion and then submit no more than three most important concerns or questions you have and how they might impact the results of the Belonging Survey.


**Write your answer here:**

1. With the concept of weighting in mind, the gender split of the overall survey stands out to me. It's not overly concerning, though the sample definitely doesn't represent the split of the UMD population.

2. Once again, keeping with the theme of survey weighting, the apendix notes that the racial breakdown of respondants, is similarly skewed and doesn't represent the general population across campus.

#-------------------------------------------------------------------------------
#Survey Weighting
#-------------------------------------------------------------------------------


```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)

```

# Introduction

This notebook explores how survey weighting affects data analysis and interpretation in journalism. We'll use a dataset from a 2020 survey on voting behavior to understand weighting concepts and how they can change our understanding of public opinion data. The codebook with the questions from the survey is here: https://github.com/dwillis/jour405_files/blob/main/nonvoters_codebook.pdf

## What is Survey Weighting?

Survey weighting is like adjusting the volume on different voices in a conversation to make sure everyone is heard equally. Think of it like running a mixing board at a concert. In survey research:

- Some groups of people are more likely to answer surveys than others
- This can make the survey sample different from the actual population
- Survey weights help fix this imbalance by giving more "volume" to underrepresented groups and less to overrepresented groups

**For example:** If your survey has too many college graduates compared to the general population, you might give each college graduate's response a weight of 0.8 (counting as 80% of a response) and each non-college graduate a weight of 1.2 (counting as 120% of a response).

## Why Should Journalists Care?

- Unweighted results might give a misleading picture of public opinion
- Proper weighting helps ensure your reporting accurately represents the community
- Understanding weighting helps you critically evaluate polls from other sources
- Small differences in weighting can sometimes change the headline of your story

## Our Questions

1. How does weighting affect our estimates of voting intentions?
2. Which demographic groups show the largest differences between weighted and unweighted results?
3. How might weighting influence reporting of this data?

# Data Overview

The nonvoters dataset contains survey responses about voting intentions and behaviors from 5,836 American adults. The survey was conducted prior to the 2020 election and includes a variety of demographic information and political attitudes.

```{r load-data}
# Load the dataset
nonvoters_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/nonvoters_data.csv")

# Look at the structure of the data
glimpse(nonvoters_data)
```

Let's break down the key variables we'll focus on:

- `weight`: The survey weight assigned to each respondent, showing how much their response "counts"
- `Q21`: Voting intention with these values:
  - 1 = Yes (plans to vote)
  - 2 = No (does not plan to vote)
  - 3 = Unsure/Undecided
- `voter_category`: Respondent's voting history:
  - "always": Votes in all or nearly all elections
  - "sporadic": Votes occasionally
  - "rarely/never": Rarely or never votes
- Demographics: `race`, `gender`, `educ` (education level), `income_cat`

# Understanding Survey Weights

Before we dive into the analysis, let's examine the weights in our dataset to understand how they're distributed.

```{r examine-weights}
# Summary statistics of weights
summary(nonvoters_data$weight)
```


```{r}
# Histogram of weights
ggplot(nonvoters_data, aes(x = weight)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "white") +
  theme_minimal() +
  labs(
    title = "Distribution of Survey Weights",
    x = "Weight",
    y = "Count"
  )
```

**Explanation:** 
- The summary statistics show us the minimum, maximum, and typical weights in the dataset
- The histogram shows how many respondents have each weight value
- Most weights cluster around 1.0 (which means those respondents count as exactly one person)
- Weights below 1.0 mean a respondent is overrepresented (counts as less than one person)
- Weights above 1.0 mean a respondent is underrepresented (counts as more than one person)

**Questions:**

1. What is the range of weights in this dataset? **Write your answer here:**
The range of weights extends from a minimum of 0.23 to a maximum of 3.04

2. What does it mean when a respondent has a weight of 0.23? What about 3.04? **Write your answer here:**
A weight of 0.23 counts their response as 23% of a response, while 3.04 is counted as 304% of a response

3. Why might some respondents be weighted more heavily than others? **Write your answer here:**
Respondents toward the 0.23 side of the data set fall into a demographic that's over represented by the data set, and thus, their response is "dimmed" slightly, so to speak. On the other end, a response with a weight closer to 3 represents a demographic that is under represented by the sample. 

# Exercise 1: Voter Categories and Weighting

First, let's look at how many people fall into each voting category in our sample.

```{r voter-categories}
library(knitr)
# Count of respondents by voter category
nonvoters_data |>
  count(voter_category) |>
  mutate(percentage = n / sum(n) * 100)
```

**Explanation:** This table shows the breakdown of our sample by voting history. We can see that most respondents (44.1%) are "sporadic" voters, followed by "always" voters (31.0%), and "rarely/never" voters (24.9%).

Now, let's compare unweighted and weighted voting intentions by voter category.

## Part A: Unweighted Analysis

First, we'll calculate the raw percentages without applying weights:

```{r unweighted-by-category}
# Calculate unweighted percentages
unweighted_by_voter_category <- nonvoters_data |>
  # Filter out missing responses
  filter(!is.na(Q21), Q21 > 0) |>
  # Group by voter category and response
  group_by(voter_category, Q21) |>
  # Count responses in each group
  summarize(count = n(), .groups = "drop_last") |>
  # Calculate percentages within each voter category
  mutate(total = sum(count),
         percentage = count / total * 100) |>
  ungroup()

# Create a more readable version with voting intentions as columns
unweighted_summary <- unweighted_by_voter_category |>
  pivot_wider(
    id_cols = voter_category,
    names_from = Q21,
    values_from = percentage,
    names_prefix = "pct_"
  ) |>
  rename(
    "Yes (%)" = pct_1,
    "No (%)" = pct_2,
    "Unsure (%)" = pct_3
  )

unweighted_summary
```

**Explanation:**
- This table shows the percentage of respondents in each voter category who said "Yes," "No," or "Unsure" when asked if they plan to vote
- We calculated these percentages by simply counting responses and dividing by the total in each category
- No weights have been applied yet—each respondent counts equally
- Notice that even among "rarely/never" voters, a high percentage say they plan to vote in the upcoming election

## Part B: Weighted Analysis

Now, let's apply the survey weights to the same analysis. Instead of just counting respondents, we'll sum their weights:

```{r weighted-by-category}
# YOUR TASK: Calculate the weighted percentages
# Use the weight variable to adjust the counts

weighted_by_voter_category <- nonvoters_data |>
  # Filter out missing responses, same as before
  filter(!is.na(Q21), Q21 > 0) |>
  # Group by voter category and response
  group_by(voter_category, Q21) |>
  # Instead of counting, sum the weights
  summarize(weighted_count = sum(weight), .groups = "drop_last") |>
  # Calculate percentages based on weighted counts
  mutate(weighted_total = sum(weighted_count),
         weighted_percentage = weighted_count / weighted_total * 100) |>
  ungroup()

weighted_summary <- weighted_by_voter_category |>
  pivot_wider(
    id_cols = voter_category,
    names_from = Q21,
    values_from = weighted_percentage,
    names_prefix = "pct_"
  ) |>
  rename(
    "Yes (%)" = pct_1,
    "No (%)" = pct_2,
    "Unsure (%)" = pct_3
  )

weighted_summary
```

**Explanation:**
- Now we've calculated percentages by summing the weights instead of simply counting people
- Each respondent contributes to the total based on their assigned weight value
- This gives us estimates that should better reflect the true population
- Notice how some percentages have changed from the unweighted analysis

## Part C: Comparing Differences

Let's create a comparison table to clearly see the differences between weighted and unweighted results:

```{r category-comparison}
comparison <- unweighted_summary |>
  # Join the two tables to compare them
  inner_join(weighted_summary, by = "voter_category", suffix = c("_unweighted", "_weighted")) |>
  # Calculate the differences in percentage points
  mutate(
    yes_diff = `Yes (%)_weighted` - `Yes (%)_unweighted`,
    no_diff = `No (%)_weighted` - `No (%)_unweighted`,
    unsure_diff = `Unsure (%)_weighted` - `Unsure (%)_unweighted`
  ) |>
  # Select just the differences for our table
  select(voter_category, yes_diff, no_diff, unsure_diff) |>
  rename(
    "Yes (% point diff)" = yes_diff,
    "No (% point diff)" = no_diff,
    "Unsure (% point diff)" = unsure_diff
  )

comparison 
```

**Explanation:**
- This table shows how many percentage points the results changed after applying weights
- Positive numbers mean the weighted percentage is higher; negative numbers mean it's lower
- For example, after weighting, the percentage of "rarely/never" voters saying "Yes" decreased by 3.6 percentage points
- These differences might seem small, but in a close election, even small shifts can be meaningful for reporting

**Answer These Questions:**

1. Which voter category shows the biggest difference between weighted and unweighted results? **Write your answer here:**
The biggest difference can be seen in the category of rarely/never voters who plan to vote in this year's election.

2. How might these differences affect reporting on likely voter turnout? **Write your answer here:**
The weighted summary underestimates voter turnout slightly, which is particularly important in this instance, because it's a group that usually doesn't vote, saying that it will vote. Generally, this will affect media coverage, because reporters won't fully grasp the anticipated turnout for a group that's somewhat of a "wild card" when it comes to voting anyways, seeing as they are typically absent. 

3. Why do you think these differences exist? What does this tell us about who might be over or under-represented in the sample? **Write your answer here:**
Like everything, weighting isn't a perfect science, so naturally some groups will usually be slightly over or under represented in a give survey. The rarely/never-Yes category was always going to be slightly odd to capture, seeing as its a group of people who rarely ever vote, contradicting their tendencies by saying that they'll do so in this election. That alone seems like it's bound to create some inconsistencies. Generally speaking, the rarely/never row features the highest variation between weighted and unweighted results, and I think some of that can be attributed to some of the unreliability on the parts of those who are answering.   
