---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2024.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

ANSWER HERE: 
The fairly extensive data set includes salaries for 564 white house employees. The position titles vary because most have highly specific roles, but generally speaking, there are numerous assistants, special assistants (many to the president), deputy assistants, policy advisors (senior and standard), and so on. Most salaries for given positions are fairly consistent. For example, the vast majority of special assistants to the president make either 139,500 a year or 121,500. From a reporting standpoint, there are only two areas that I think may skew the data set, and that’s naturally the top and bottom. 11 employees have a salary of 0 for whatever reason, and more notably, Associate Counselmember Michelle M. Barrans makes 251,258 per year, which I thought was a little peculiar because no one else in the salary list makes over 200,000. Because of how many people are included in the data set, those outliers probably don't alter the mean or median too much, but I still find them kind of odd.     

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |>
  summarize(mean_salary = mean(salary),
          median_salary = median(salary)  
            )
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |>
  summarize(mean_salary = mean(salary),
          median_salary = median(salary),
          min_salary = min(salary),
          max_salary = max(salary)
            )

```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

ANSWER HERE: 
On the surface, nothing about the data appears incredibly out of the ordinary. The mean and median both fall generally in line with what one would expect from White House salaries, though the median salary is greater than the mean. That left-skewed data would indicate that the average White House staffer gets paid slightly more than the mean suggests — but again, the difference isn’t staggering. If anything, the difference is probably due to the 11 staffers who have salaries of 0, which negatively skew the mean. As I mentioned earlier, the maximum value isn’t too noteworthy on its own, but with the contest that the next highest paid employee receives almost 60,000 less, I think it could be significant. 

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

ANSWER HERE:

While the 11 reportedly unpaid staffers feel like it has story potential, I feel like there’s probably a decent explanation for it (intern, volunteer positions, they don’t have the exact salary number for whatever reason). With that being said, I think it would be more worthwhile to look into Michelle M. Barrans’ role in the white house, and what she does as an Associate Counsel that allows her to be the highest-paid staffer by a considerable margin. Several other Associate Counsel’s get compensated fairly well, but none as much as her. My next step would be researching her background and time in the White House to see if I can find anything about her role that would indicate why she’s so highly paid. I think it could also be helpful to talk to other Associate Counsel’s to learn more about her, since I imagine they work fairly closely together. Once I have that background, I would try to make contact with her in any way to get her direct perspective on her role in the White House. Overall, I think it could turn into an investigative piece or even an interesting profile if there’s a good reason why she gets paid so much more than everyone else.  
