---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities |> 
  mutate(violent_rate_2019 = violent2019/pop2019 * 1000,
         property_rate_2019 = property2019/pop2019 * 1000,
         violent_rate_2020 = violent2020/pop2020 * 1000,
         property_rate_2020 = property2020/pop2020 * 1000
  
)
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

ANSWER HERE: 
There are several notable trends in this data set, but starting on the positive end, there were a handful of cities/towns that didn't log any violent or property crimes in either year; almost all had populations below 1,000, with no town having more than 2,000 people in either year. This fact isn't overly surprising, because places with low populations don't typically see high crime rates. Generally, higher crime rates were seen in municipalities with greater populations. 

Ocean City was the primary example of high crime rates across the two-year study. Of the four categories being tracked, the only one that didn't have Ocean City at the top was the violent crime rate in 2019 (Ocean City was second to Baltimore). Property crime was a particular issue for Ocean City during both years. The gap (by rate) between Ocean City and the next closest city was sizeable in both 2019 and 2020.  

I also found the year-to-year change (or lack thereof) to be significant. For example, of the 10 cities with the highest 2019 property crime rates, seven were in the top 10 in 2020, too. As for violent crime rates across the two years, eight cities were among the top-10 highest in the state in both 2019 and 2020.

### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates |>
  mutate(pct_change_property = (property_rate_2020 - property_rate_2019)/property_rate_2019 * 100,
         pct_change_violent = (violent_rate_2020 - violent_rate_2019)/violent_rate_2019 * 100
         )
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

ANSWER HERE:
Starting with the outliers, for property rate changes, there are four cities that saw an infinite jump in reported property crimes (because there were zero reported in 2019). Still, none of those seemed overly noteworthy or significant to me since that occurred in relatively small communities. The rise in total property crime from one year to the next in all of those places was relatively small. On the other end, five towns saw a 100% decrease in property crime rate (for the opposite reason as the four cities that saw an infinite increase), and a decent number of towns also saw "NaN" (zero property crimes reported across either year). With that in mind, I thought the most notable decrease in property crime rate from 2019 to 2020 was seen in Fredrick -- a relatively sizeable population center that saw a 90% decrease in property crime rate from 2019 to 2020. 

A really similar principle applies to the change in violent crime rate, too. There's a handful of places that saw an infinite increase, a 100% decrease, or NaN -- none of which particularly stood out to me because all of those occurred in pretty small towns, where outliers like those are understandably common. But once again, I noticed that Frederick saw another gargantuan decrease in violent crime rates, with a -95% change from 2019 to 2020. Those types of drops in a place with over 70,000 people seem almost impossible, but assuming their figures are accurate, I think there could be a really good story done to explore how Frederick reduced overall crime so drastically from one year to the next.

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

ANSWER HERE:

Baltimore is often profiled as one of the most crime-ridden cities in the nation. Recent data suggests it might not even be the most troubled area in its own state.

In 2019, the eastern shore vacation community of Ocean City recorded around 103 property crimes per 1000 people. That figure actually increased to around 121 in 2020. No other municipality reported over 100 property crimes per 1000 residents in either year. Ocean City also saw similarly high violent crime rates across the two-year study, highly concerning figures for a local getaway destination.


### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
