---
title: "HW7_MandMs"
date: 10/1/2025
---


In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load an M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv
```{r}
M_and_Ms2 <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```


### Task 3: Sample 5 rows of data from the dataset, and then sample 10 rows, saving each to separate dataframes called sample5 and sample10. Be sure to use set.seed(1234) before you start.
```{r}
set.seed(1234)

sample_5 <- M_and_Ms2 |>
  sample_n(5)

sample_10 <- M_and_Ms2 |>
  sample_n(10)

```


### Task 4: Calculate the mean, median and standard deviation of red M&Ms from each of the samples.
```{r}
sample_5 |> summarize (mean = mean(red), median = median(red), sd = sd(red))

sample_10 |> summarize (mean = mean(red), median = median(red), sd = sd(red))
```


### Task 5: Calculate the mean, median and standard deviation of red M&Ms from the population

```{r}
M_and_Ms2 |> summarize (mean = mean(red), median = median(red), sd = sd(red))
```

### Task 6: Make three histograms: one each for sample5, sample10 and the original dataset. Include a headline. Use a consistent binwidth.

```{r}

M_and_Ms2 |> 
  ggplot(aes(x = red)) + 
  geom_histogram(binwidth = 2, fill = "blue", color = "black") +
  labs(title = "Distribution of Red M&Ms in the Total Population", x ="Red M&Ms Count", y ="Frequency")

sample_10 |>
  ggplot(aes(x = red)) + 
  geom_histogram(binwidth = 2, fill = "blue", color = "black") +
  labs(title = "Distribution of Red M&Ms in the Sample of 10", x ="Red M&Ms Count", y ="Frequency")

sample_5 |>
  ggplot(aes(x = red)) + 
  geom_histogram(binwidth = 2, fill = "blue", color = "black") +
  labs(title = "Distribution of Red M&Ms in the Sample of 5", x ="Red M&Ms Count", y ="Frequency")

```


### Task 7: Answer questions


Compare the three sets of means, standard deviations and histograms. Which are more helpful to you in understanding the accuracy of the samples?

ANSWER: The contrast between the means and histograms makes this question kind of difficult to answer. On one hand, the sample5 produced a mean of 73.8 and the sample10 produced a mean of 74.7 -- both of which were very similar to the population mean of 72.98. Looking at the mean in a vacuum, I'd say I got very accurate samples, but the histograms would suggest otherwise. The total population histogram is right-skewed with a large number of values spread between 60 and 80. The sample10 histogram is generally the same, but looks funkier because of the two +80 and one +90 values. The sample5 histogram is even more unorthodox, with a lot of cavernous space between 60 and 80, but the numbers are evidently balanced out by the three relatively extreme values. With all of that being said, I think the histograms give a slightly more accurate representation of the samples, since I can see the actual distribution of values. Overall, the samples are both fairly accurate, but I also think I got relatively "lucky" with my samples.

What does this exercise tell you about the relationship between sample size and estimate accuracy?

ANSWER: The results of this exercise actually suggest that sample size isn't a huge factor in estimate accuracy. From a mean standpoint, my sample5 was actually a slightly more accurate depiction of the overall data set than my sample10. Both sample means were honestly very close to the overall mean, too, despite both samples being very small portions of the full data set, which included 100 observations. Despite what this instance showed, though, logical thinking suggests that a larger sample usually gives a more accurate representation of a total population, assuming there's no sampling bias.

How does random sampling introduce variation into statistical estimates?

ANSWER: Random sampling (especially when the sample is so small in comparison to the overall population) could introduce extreme variation into statistical estimates, especially for data sets with high standard deviations. For example, a random sample of 10 from this M&M data set could theoretically turn up with 10 counts that all had 97 or more red M&Ms, which would be an extremely inaccurate representation of the data. But, a random sample that accounts for somewhere closer to 40 or 50 percent of the population is usually bound to be more accurate than a same at 5 or 10 percent.

If you repeated this process, would the samples always be identical? Why or why not?

ANSWER: No, the samples would likely be different because that's the point of a random sample -- it's all happen chance. In theory, if I pulled another sample of five, I could get five examples of observations that saw less than 50 red M&Ms, and equally, I could get five examples of observations that saw more than 90 red M&Ms. Overall though, I'd be thoroughly shocked to get sample means so close to the population mean, if I were to run teh experiment again with a sample of 5 and sample of 10. 


When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
