
---
title: "Activity: Linear Regression "
date: 11-17-2025
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

## Identifying Schools with Biggest Differences in Math Test Scores

### Introduction

In this analysis, we will create a scatterplot to visualize the relationship between the FARMS (Free and Reduced-price Meal System) percentage and the percentage of students who met or exceeded math expectations. We'll then use linear regression to identify the schools that show the biggest differences between their actual and expected math test scores.

This analysis can be useful for journalists to understand which schools are performing better or worse than expected, based on the socioeconomic status of their student population. By identifying these schools, targeted interventions and support can be provided to help improve student outcomes - and to provide some context on comments by policymakers.

### Loading the Data

First, let's load the data into a data frame using the read_csv() function from the tidyverse library.

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_pa.csv")
```
### Task: Create a Scatterplot and Describe the Results (2 points)

Next, let's create a scatterplot to visualize the relationship between the FARMS  (Free and Reduced-price Meal System) percentage and the percentage of students who met or exceeded math expectations. We'll use ggplot2 to create the plot and add a linear regression line.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations", title = "Scatterplot of Math Expectations vs FARMS Percentage")
```

The scatterplot shows the overall trend of the relationship between the FARMS percentage and the math test scores. The red line represents the linear regression line, which helps us understand the average relationship between the two variables. Based on what you see here, what's the relationship between FARMS percentage and math test scores?

**WRITE YOUR ANSWER HERE**
Based on the chart above, ther is no clear relationship between FARMS percentage and math test scores. Very few dots within the scatter plot fall near the line of best fit, which, in this instance, hardly captures the incredibly varied sirtibution of the data. 

### Task: Linear Regression Model (3 points)

Next, we'll perform a linear regression analysis using the lm() function. We'll set the percentage of students who met or exceeded math expectations as the dependent variable, and the FARMS percentage as the independent variable. In other words, if we know the FARMS percentage for a school, how reliably can we predict its math score?

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)

summary(model)
```

There are three things we need to examine here: the p-value, the Adjusted R-squared value and the coefficients. You know what a p-value is and how to evaluate it. The Adjusted R-squared is a measure of how much of the difference between schools' math scores can be explained by their FARMS percentage. You don't need to multiply that figure by anything: that's a percentage. The closer the adjusted R-squared is to 100, the more of the outcome the independent variable explains. If both the p-value and adjusted R-squared values are significant, we can look to the coefficients.

In the middle, you can see the (Intercept) is 59.57897 and the FARMS coefficient is 0.12349. Remember high school algebra? Remember learning the equation of a line? Remember swearing that learning y=mx+b is stupid because you’ll never need it again? Surprise. It’s useful again. In this case, we could try to predict a school's score math score by using y=mx+b. In this case, y is the math score, m is 0.12349 and b is 59.57897. So we would multiply a school's FARMS percentage by 0.12349 and then add 59.57897 to it. The result would tell you what the math score would be, according to our model.

Based on what you've learned so far, is this model worth using? What does it tell us about the nature of the relationship between FARMS percentage and math scores?

**WRITE ANSWER HERE**

This model seems to be worth using because it proves that there's no significant connection between the two variables. Harkening back to the y=mx+b example, if you were to enter the FARM score for the vast majority of the data points, the corresponding y value would produce a value on the line of best fit. There would be a large discrepency between most of the scatter plot points, and said value on the line of best fit. That said, the equation can be used to very accurately predict a select few of the values in the data set.   

### Task: Identifying Schools with Biggest Residuals (2 points)

To find the schools with the biggest differences between their actual and expected math test scores, we'll extract the residuals from the linear regression model. The residuals represent the difference between the actual and predicted values. Residuals can tell you several things, but most important is if a linear model the right model for your data. If the residuals appear to be random, then a linear model is appropriate. If they have a pattern, it means something else is going on in your data and a linear model isn’t appropriate.

Residuals can also tell you who is under-performing and over-performing the model. And the more robust the model – the better your r-squared value is – the more meaningful that label of under or over-performing is. 

We'll then use the arrange() function to sort the data frame by the absolute value of the residuals.

```{r}
schools$predicted <- predict(model)

# Calculate the residuals
schools <- schools |> 
  mutate(residual = `% Met or Exceeded Math Expectations` - predicted)

# Sort the data frame by the absolute value of the residuals in descending order
schools <- schools |>
  mutate(abs_residual = abs(residual)) |>
  arrange(desc(abs_residual))

```
The schools with the biggest positive residuals are performing better than expected, while the schools with the biggest negative residuals are performing worse than expected, based on their FARMS percentage. Look at the absolute values of the residuals; what are the largest differences between the predicted result and the actual math scores?

**WRITE YOUR ANSWER HERE**
The following schools are over performing significantly: Crooked Billet Elementary School, Woodland Elementary School, and Upper Moreland Elementary School --> all three are over or around the +25 mark. On the other end of the spectrum, Whitemarsh Elementary School, Gulph Elementary School, and Lower Gwynedd Elementary School are all massively under performing.

## Task: Evaluate the model (3 points)

A school board member has said at a meeting that elementary schools with fewer students on free and reduced price meals do better on standardized math tests. Describe what conclusions you'd draw from this model's results and how it would influence a story pitch about the school board member and math scores.

**WRITE YOUR ANSWER HERE**

Based on this model, there's a very weak correlation between the FARMS program and math scores (schools that have higher participation in the farms program don't necessarily have better math scores and vice versa). I think the fact that the school board member made a pretty faulty claim is worth a story as is. As for a story on the topic though, I think it's worth talking to the school board member to find out where they are sourcing their information from. Having the board member's perspective in the story feels highly important from a standpoint of highlighting both sides.  
