---
title: "Oct 1 Activities: M&Ms; Survey; Sampling"
date: 9-30-2025
---

All three class activities are on this .rmd notebook

Activity #1: Activity: M&Ms - Central Limit Theorem

Working in groups, each group will get a handful of candy from the bag, and then count the number of red and black candies. Log those in this Google Sheet 

https://docs.google.com/spreadsheets/d/1KOeoVTq2W7L2T9sWjwh5U_2f59GsTeAylVGD53rNWww/edit?usp=sharing

(give your group a nice name).

Take a look at the values that we all get - what do you notice about the patterns? This is an exercise about the central limit theorem. 

The central limit theorem states that the sampling distribution of the mean of any independent, random variable will be normal or nearly normal, if the sample size is large enough.

Then discuss and try to answer these questions:

--How close is your group's mean to the true proportion of 0.5?
--Why do different groups get different results?
--What happens to individual extreme results when we combine all groups?
--How does this relate to polls and surveys in real life?

Write your answer below. And then you may eat the M&Ms.

**A:**
Question 1: Our group's mean is 90% to true proportion of 0.5

Question 2: Every bag of M&M's is sorted differently, menaing that there's a different number of individual colors in each bag. 

Question 3: Those extreme results would likely level out, making the color distribution more normal if all groups in the class were to be combined.

Question 4: The greater the sample size, the more likely a data set is to be effected by outliers or extraneous data points. When doing political polls, or other surveys, having a larger sample size (in a random sample) typically helps us get a more accurate understanding of public perception.  
------------------------------------------------------------------------------

# Activity #2:

Belonging Survey

Working in pairs, take a look back at the main findings from UMD's Belonging Survey:

https://president.umd.edu/internal/belonging


Specifically, focus on the "Sample Description" section and the Appendix. Discuss the differences and between the population and the sample, and what the document says about how the survey attempted to deal with them. Are there any decisions described in the document that raise questions in your minds? Detail the differences you see and your understanding of the methods they employed (it's ok if you don't understand all of them), and add your questions. You should discuss this as a pair but submit individually.

**A:**
The piece that first stuck out to us was the fact that the sample includes 23% of the population, which sounds like a small portion, but equally, the population is incredibly, with over 53,000 people. So, with that in mind, the 23% feels fairly impressive. Contrasting that though, the population also doesn't seem to be equally distributed by gender groups, as women and gender expansive people were over represented, while men were underrepresented. Similarly, Asian/Asian American and multiracial students were slightly over represented, while Hispanic students were underrepresented.        
------------------------------------------------------------------------------

# TASK 3: SAMPLING

Based on the class sampling demo (available on Elms), sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

**A**
```{r}
library(tidyverse)

```


### Task 2: Load 1,000 rows of data with the ages of people from 18-80

**A**
```{r}
set.seed
younguns <- tibble(
  age = sample(18:80, 1000, replace = TRUE)
)

head(younguns)
```

### Task 3: Sample 100 rows of data from the dataset
**A**
```{r}
sample_100 <- younguns |>
  sample_n(100)

sample_50 <- younguns |>
  sample_n(50)

```


### Task 4: Calculate the mean age and standard deviation of the sample
**A**
```{r}
sample_100 |> summarize (mean = mean(age), median = median(age), sd = sd(age))

sample_50 |> summarize (mean = mean(age), median = median(age), sd = sd(age))

```


### Task 5: Calculate the mean age and standard deviation of the population
**A**
```{r}
younguns |> summarize (mean = mean(age), median = median(age), sd = sd(age))

```


### Task 6: Compare the sample and population statistics

Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?

**A**: My sample (which only accounts for 10% of the total population) data is significantly different from the numbers produced from my total population figures. The younguns data set had a mean age of 21.1 and a median age of 21.6, with a standard deviation of 6 -- an indication of a fairly even distribution. The mean and median were similar, close together from my sample of 100, though the figures were higher, at a mean of 49.6 and a median of 50. A sample of 50 returns similar results, with a mean of 49.2 and a median of 49. Both standard deviations from the samples are nearly three times higher than that of the younguns total population. Overall, the differences between the samples and the total population prove that a random sample of 10% or lower simply can't produce an accurate representation of the full data set.


When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
